### 2024-04-09 Chapters/Chapter1.tex
Aðeins uppfært eftir ath frá KJ 9/4/24

%----------------------------------------------------------------------------------------
Wind gusts are brief increase in wind speed (lasting seconds) as compared to mean wind speed. The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period. The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute interval. This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}. As the Navier Stokes Equation (\ref{eq:navierstokes}) shows the variability of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth, and resistance. It states that the change in wind is described by the pressure gradient, the oscillation, the acceleration due to gravity and the resistance from the landscape.
\begin{equation}
    \label{eq:navierstokes}
    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
\end{equation}
Equation \ref{eq:navierstokes} relates the change in wind w.r.t. time ($\frac{\delta \boldsymbol{V}}{\delta t}$) and change in wind w.r.t. position ($\nabla\boldsymbol{V}$) to pressure change w.r.t. position ($\nabla\boldsymbol{P}$). The wind is also influenced by the oscillation, gravity and resistance but the main drive of wind is pressure gradient\cite{uncertainties_in_numerical_weather_predictions}.

@@

\section{Background}
The history of numerical weather predictions goes all the way back to the 1920's when Lewis Fry Richardson pioneered the field and tried to produce forecasts. The results were flawed and impractical because of the time it took to calculate was much longer than the forecast period. In the 1950's, with the advent of computers we get the first operational forecasts. In September of 1954, Rossby and his Stockholm based team produced the first real-time barotropic forecasts. The next year the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton New Jersey, released their first forecasts. These forecasts were for 36 hours at 400, 700 and 900 mb. The results were inferior to subjective human-based forecasts but showed that such forecasts were feasible and promoted further development in the area \cite{historyNWP}. The field underwent significant transformation over the next 50 years, and predictions got significantly better.

@@

In the last two years there have been even more developments with the emergence of Large AI Weather forecast Models (LWM). In 2024 Ling et al. tried to standardize the definition of LWM in meteorology and came up with 3 rules that need to be met to count as LWM.

\begin{enumerate}[label = Rule \arabic*:]
    \item Large Parameter Count. The number of parameters can vary wildly but a general range might be from tens of millions to billions of parameters
    \item Large Number of Predictands: predicting on different levels (such as pressure levels or height levels) and offering detailed information on the atmospheric vertical structure and surface conditions
    \item Scalability and downstream applicability. This might crystallize in predicting cyclones. Often, the teams responsible for creating these models try to show their applicability to predict cyclones when not trained specifically on cyclone data (e.g.  \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast})\cite{SecondRevolution}. This is done to show the versability of the models.
\end{enumerate}
Before 2022, LWM had been shown to be able to compete with traditional NWP for some specifc cases as well as making predictions quicker, after training. No model had shown that it could in any way completely replace the traditional systems. In early 2022, Pathak et al. presented FourCastNet. FourCastNet uses an Adaptive Fourier Neural Operator model that leverages transfomer architecture rather than the popular convolutional model architechture. FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms for smaller variables. It generates a week-long forecast in less than 2 seconds, orders of magnitude faster than standard physical methods\cite{FourCastNet}. In 2022, machine learning methods were presented that made predictions much faster than traditional NWP, after a one time training (or at least training that wouldn't have to be redone often). These were in some cases performing better than NWP.  In 2023, Remi Lam and the GraphCast team at Google introduced GraphCast. This model was able to outperform the industry standard High Resolution Forecast (HRES) produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This model as the name suggests leverages graphing connections rather than traditional grid like data structure. The base data is given in latitude and longitude degrees at a resolution of 0.25 degrees. This means points are closer to each other at the poles. Using the graphing structure is supposed to help with bias incurred as a result of this\cite{GraphCast}.

There has been a lot of progress made over the last 6 years (since 2018) and especially in the last 2 years (since 2022). The progression from machine learning methods being an interesting idea in the field of numerical weather predictions, to outperforming the standard NWP has been remarkably quick. Two years ago, machine learning methods were able to predict quickly and in some niche cases outperform traditional models. They were not generally competitive with standard weather models. Now they are competitive. It will be very interesting to watch what the next few years will have in store for the development of machine learning in weather predictions.

@@

In 2004, H.Ágústsson and H. Ólafsson looked at the variability of gust factor in complex landscapes. They looked at data from automatic weather stations that measure wind at 10 meters above ground. A common definition of wind gust is the maximum average 1 second measured wind speed over a rolling 3 second interval using all measurements over a ten minute span. A common definition of average wind is the average of the measured wind speed over 10 minutes. These are the definitions used both by H. Ágústsson and H. Ólafsson in 2004 as well as this thesis. The data that was studied in 2004 comes from the same source as used in this thesis, but limits itself to a smaller section. They only looked at the years 1999-2001. They looked at three factors and how these three parameters effected the gust factor. These were $d_m, D, H$, that is direction of wind blowing off a mountain, distance to the mountain and the height of the mountain above the weather station. Their main results were these, gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain. The study in 2004 looked at the effect of a dominant point upwind. It did not look at the effects of the landscape more broadly. In this study, landscape upwind is looked at along with landscape downwind. Both the upwind and downwind landscape can influence the gustiness at a certain point \cite{GNP_vidtal}.

>>>>

%----------------------------------------------------------------------------------------
Wind gusts are brief increase in wind speed (lasting seconds) as compared to mean wind speed. The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period. The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute interval. This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}. As the Navier Stokes Equation (\ref{eq:navierstokes}) shows the variability of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth, and resistance. It states that the change in wind is described by the pressure gradient, the oscillation, the acceleration due to gravity and the resistance from the landscape.
\begin{equation}
    \label{eq:navierstokes}
    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
\end{equation}
Equation (\ref{eq:navierstokes}) relates the change in wind w.r.t. time ($\frac{\delta \boldsymbol{V}}{\delta t}$) and change in wind w.r.t. position ($\nabla\boldsymbol{V}$) to pressure change w.r.t. position ($\nabla\boldsymbol{P}$). The wind is also influenced by the oscillation, gravity and resistance but the main drive of wind is pressure gradient\cite{uncertainties_in_numerical_weather_predictions}.

@@

\section{Background}
The history of numerical weather predictions goes all the way back to the 1920's when Lewis Fry Richardson pioneered the field and tried to produce forecasts. The results were flawed and impractical because of the time it took to calculate was much longer than the forecast period. In the 1950's, with the advent of computers the first operational forecasts emerged. In September of 1954, Rossby and his Stockholm based team produced the first real-time barotropic forecasts. The next year the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton New Jersey, released their first forecasts. These forecasts were for 36 hours at 400, 700 and 900 mb. The results were inferior to subjective human-based forecasts but showed that such forecasts were feasible and promoted further development in the area \cite{historyNWP}. The field underwent significant transformation over the next 50 years, and predictions got significantly better.

@@

In the last two years there have been even more developments with the emergence of Large AI Weather forecast Models (LWM). In 2024, Ling et al.\cite{SecondRevolution} tried to standardize the definition of LWM in meteorology and came up with 3 rules that need to be met to count as LWM.

\begin{enumerate}[label = Rule \arabic*:]
    \item Large Parameter Count. The number of parameters can vary wildly but a general range might be from tens of millions to billions of parameters
    \item Large Number of Predictands: predicting on different levels (such as pressure levels or height levels) and offering detailed information on the atmospheric vertical structure and surface conditions
    \item Scalability and downstream applicability. This might crystallize in predicting cyclones. Often, the teams responsible for creating these models try to show their applicability to predict cyclones when not trained specifically on cyclone data (e.g.  \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast})\cite{SecondRevolution}. This is done to show the versability of the models.
\end{enumerate}
Before 2022, LWM had been shown to be able to compete with traditional NWP for some specifc cases as well as making predictions quicker, after training. No model had shown that it could in any way completely replace the traditional systems. In early 2022, Pathak et al.\cite{FourCastNet} presented FourCastNet. FourCastNet uses an Adaptive Fourier Neural Operator model that leverages transfomer architecture rather than the popular convolutional model architechture. FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms for smaller variables. It generates a week-long forecast in less than 2 seconds, orders of magnitude faster than standard physical methods\cite{FourCastNet}. In 2022, machine learning methods were presented that made predictions much faster than traditional NWP, after a one time training (or at least training that wouldn't have to be redone often). These were in some cases performing better than NWP.  In 2023, Remi Lam and the GraphCast team at Google introduced GraphCast. This model was able to outperform the industry standard High Resolution Forecast (HRES) produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This model as the name suggests leverages graphing connections rather than traditional grid like data structure. The base data is given in latitude and longitude degrees at a resolution of 0.25 degrees. This means points are closer to each other at the poles. Using the graphing structure is supposed to help with bias incurred as a result of this\cite{GraphCast}.

There has been a lot of progress made over the last 6 years (since 2018) and especially in the last 2 years (since 2022)\cite{SecondRevolution}. The progression from machine learning methods being an interesting idea in the field of numerical weather predictions, to outperforming the standard NWP has been remarkably quick. Two years ago, machine learning methods were able to predict quickly and in some niche cases outperform traditional models. They were not generally competitive with standard weather models. Now they are competitive. It will be very interesting to watch what the next few years will have in store for the development of machine learning in weather predictions.

@@

In 2004, H. Ágústsson and H. Ólafsson\cite{mean_gust_HA_HO} looked at the variability of gust factor in complex landscapes. They looked at data from automatic weather stations that measure wind at 10 meters above ground. A common definition of wind gust is the maximum average 1 second measured wind speed over a rolling 3 second interval using all measurements over a ten minute span. A common definition of average wind is the average of the measured wind speed over 10 minutes. These are the definitions used both by H. Ágústsson and H. Ólafsson in 2004 as well as this thesis. The data that was studied in 2004 comes from the same source as used in this thesis, but limits itself to a smaller section. They only looked at the years 1999-2001. They looked at three factors and how these three parameters effected the gust factor. These were $d_m, D, H$, that is direction of wind blowing off a mountain, distance to the mountain and the height of the mountain above the weather station. Their main results were these, gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain. The study in 2004 looked at the effect of a dominant point upwind. It did not look at the effects of the landscape more broadly. In this study, landscape upwind is looked at along with landscape downwind. Both the upwind and downwind landscape can influence the gustiness at a certain point \cite{GNP_vidtal}.

### 2024-04-09 Chapters/Chapter2.tex
Aðeins uppfært eftir ath frá KJ 9/4/24

Data was sourced from several streams. Veðurstofan provided recordings from weather stations all around the Iceland. Data was downloaded from Copernicus Arctic Regional Reanalysis dataset (CARRA). A land elevation model was also provided by Veðurstofan.

@@

Veðurstofan provided recordings from weather stations all around Iceland. These automatic weather stations (AWS) are at a height of 10 meters above ground level. The information that is provided by these AWS is presented in two different type of documents, hourly and 10 minute documents. The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These features for each of the stations go back at least two decades, but do not have the same start date. This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\section{Carra Data}
The CARRA dataset goes back to September 1990 and is currently updated monthly, with a latency of 2-3 months. The oldest Vedurstofu data example that fulfills given criteria is from \startDateVedur. This is covered by Carra. A few of the newer points from AWS were not available from Carra when data was gathered. The Carra dataset is available for two regions, west and east. Each of these covers a vastly larger area than the area of interest. It is possible to request a subsection of the area, a rectangular subsection of the area. This leads to having to store a large amount of data, because it is not possible to ask for specific points of interest. To get the data one has two options. Their web interface or using their API client. Using the API client is the only realistic option here, as there were thousands of requests made for different times.

Each request is made by going through all instances of the Veðurstofu data were able to fulfill the requirements (of mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval). For each such observation in the Veðurstofa data, two API calls excluding redundancy are needed. One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times. That is if the observation was at 13:00, we would need Carra data for noon and 15:00. Date and times were generated automatically from the AWS and requested by calling the client. Using these datapoints, interpolation was used to get an estimation for the point of the given weather station. The Carra data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this observation was downloaded from height levels. That is, data was requested at heights of 15, 150, 250 and 500 meters above ground. For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be interpolated to create data for model to be trained on.

After using this method to request terabytes of data, it was discovered that it is apparently possible to query a specific area. This decreased the size of each file created after a request from around 50 MB to around 2 MB. Downloading and processing this data would take a lot less time. The bottleneck for retrieving the data is the request time. That is the time the request is queued and running (server preparing the data) before it begins downloading. This could range from almost immediate to 30 minutes. This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during the winter and spring of 2023/2024, which increased the wait time and sometimes resulted in queries not being responded to. This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project. Fortunately these problems were only particularly time consuming during mid winter.

\section{Elevation data}
Veðurstofa provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid. This file encompasses Iceland and is around 685 Mb. A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area. The country was divided into 10 parts (as only around 13\% of the file was able to be read into memory at each time as a part of dictionary object) with boundary boxes. A quadtree was constructed for each of the sections, so as to simplify the lookup. A quadtree allows for efficient lookup by searching for points that intersect a given boundary. A quadtree, like the name indicates, is a tree structure. To initialize it a boundary box is given. This boundary box is then the parent of exactly four nodes and so on, dividing the area in four at each level and allowing quick lookup. This was not necessary as the Python package Rasterio allowed for quick lookup with it's index and the affine transform. Using this package it is possible to quickly look up elevation given coordinates using matrix calculations.

>>>>

Data was sourced from several streams. The Icelandic Meterological Office (IMO) provided measurements from weather stations all around the Iceland. NWP data was downloaded from Copernicus Arctic Regional Reanalysis dataset (CARRA). A land elevation model was also provided by IMO.

@@

IMO provided 10 minute measurements from \nStationsMin weather stations all around Iceland. The measurements that met the filtering criteria, started in \startDateVedur and ended in 2023 (the year measurements were provided). Of these \nStationsMin stations, \nVedurMin were from IMO Íslands and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 7.5 meters above ground. The location of these weather stations can be seen in Figure \ref{fig:aws_map}. The information that is provided by these Automatic Weather Stations (AWS) is presented in two different type of data files, hourly and 10 minute files. The hourly files are summations of the 10 minute files, with the exception that errors, such as nails, still in the 10 minute files should have been removed from the hourly documents. Nails, are sharp increases from the rest of the data and are unrealistic outliers that are considered measurement errors and are discarded. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These measurements were started in the end of the 20th century, when the first AWS stations were installed. More have been added in the over two decades since then. This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\begin{figure}
    \includegraphics[scale = 0.75]{Figures/weather_stations_2024-04-09.png}
    \caption{Locations of automatic weather stations in Iceland}
    \label{fig:aws_map}
\end{figure}

These 10 minute measurements were filtered to only include measurements where the average wind speed was at or above \averageWindSpeedLimit m/s. They were also filtered in such a way that only the highest value in any given 48 hour span is considered. This can be seen in Code Listing \ref{code:filter48hr}. Any measurement that is within a 48 hour interval centered on of the current highest is removed. This is done for each station and iteratively until all measurements have been removed.

\begin{lstlisting}[style = Python, caption = {Filter points over 48 hour interval}, label = code:filter48hr]
    for station in tqdm(stations, total = len(stations)):
        subset_df = vedur_df[station == vedur_df.stod]
        subset_df = subset_df.reset_index(drop = True)

        while not subset_df.empty:
            idx = subset_df.f.idxmax()
            time_of_max = subset_df.iloc[idx].timi

            filtered_data.append(subset_df.iloc[idx])

            subset_df = subset_df[abs(subset_df.timi 
                - time_of_max) >= pd.Timedelta(threshold)]

            subset_df = subset_df.reset_index(drop = True)  
\end{lstlisting}

When these two filterings have been done, the end result is \nStationsMin stations that meet these criteria. Of these, \nVedurMin belong to IMO and \nVGMin belong to IRCA. The number of observations by year can be seen in Figure \ref{fig:occurrences_by_year}. The first observations occurring in \startDateVedur and the latest in 2023.

\begin{figure}
    \begin{center}
        \includegraphics[scale = 0.75]{Figures/occurrences_by_year.png}
    \end{center}
    \caption{Measurements that fit filtering by year (from \startDateVedur to 2023)}
    \label{fig:occurrences_by_year}
\end{figure}

Looking at the number of measurements that meet these criteria for a given station, Figure \ref{fig:occurrences_by_station_20} shows only one station has over 1000 measurements that fit the criterias.

\begin{figure}
    \begin{center}
        \includegraphics[scale = 0.75]{Figures/station_occurrences_20.png}        
    \end{center}
    \caption{Number of stations that have measurements in a range}
    \label{fig:occurrences_by_station_20}
\end{figure}

\section{CARRA Data}
The CARRA dataset goes back to September 1991 and is currently updated monthly, with a latency of 2-3 months\cite{carra_information}. The oldest IMO data point that fulfills given criteria is from \startDateVedur. This is covered by CARRA. A few of the newer points from AWS were not available from CARRA when data was gathered. The CARRA dataset is available for two regions, west and east. Each of these covers a vastly larger area than the area of interest. This leads to having to store a large amount of data, because it is not possible to ask for specific points of interest. To get the data one has two options. Their web interface or using their API client. Using the API client is the only realistic option here, as there were thousands of requests made for different times.

Each request is made by going through all instances of the IMO data after filtering (by mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval). For each such observation in the IMO data, two API calls excluding redundancy are needed. One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times. That is if the observation was at 13:00, CARRA data for noon and 15:00 would have to be queried. Date and times were generated automatically from the AWS and requested by calling the client. Using these datapoints, interpolation was used to get an estimation for the point of the given weather station. The CARRA data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this observation was downloaded from height levels. That is, data was requested at heights of 15, 150, 250 and 500 meters above ground. For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be interpolated to create data for model to be trained on.

After using this method to request terabytes of data, it was discovered that it was possible to query a specific area. This decreased the size of each file from around 50 MB to around 2 MB. The bottleneck for retrieving the data is the request time. The time the request is queued and running (server preparing the data) before it began downloading. This could range from almost immediate to 30 minutes. This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during \href{https://forum.ecmwf.int/t/a-new-cds-soon-to-be-launched-expect-some-disruptions/1607}{the winter and spring of 2023/2024}, which increased the wait time and sometimes resulted in queries not being responded to. This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project. Fortunately these problems were only particularly time consuming during mid winter.

\section{Elevation data}
IMO provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid. This file encompasses Iceland and is around 685 MB. A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area. The country was divided into 10 parts (as only around 13\% of the file was able to be read into memory at each time as a part of dictionary object) with boundary boxes. A quadtree was constructed for each of the sections, so as to simplify the lookup. A quadtree allows for efficient lookup by searching for points that intersect a given boundary. A quadtree, like the name indicates, is a tree structure. To initialize it a boundary box is given. This boundary box is then the parent of exactly four nodes and so on, dividing the area in four at each level and allowing quick lookup. This was not necessary as the Python package Rasterio allowed for quick lookup with it's index and the affine transform. Using this package it is possible to quickly look up elevation given coordinates using matrix calculations.

### 2024-04-09 Chapters/Chapter3.tex
Aðeins uppfært eftir ath frá KJ 9/4/24

This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models. When working with hundred of thousands of rows, the efficiency of the code is very important. Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible. The three data sources were all in a different format. Measurement data from Veðurstofa was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format. To use the data to train, these three data sources needed to be combined into one file. This was done based on the measurement data from the Veðurstofa. A limit was set on the average wind speed and it was used to select measurement points. Along with the average wind speed having to be above a certain limit, to make sure that the same weather for the same location is not duplicated, only the top wind speed in any given 48 hour period is selected. That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times. The data from Veðurstofan was supplied for 10 minute increments, while CARRA data is in 3 hour intervals. This means that to use the CARRA data to predict the measured values from Veðurstofa, temporal interpolation would need to be done. Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations. The elevation information was given by a 20 by 20 rectangular grid that covers Iceland. When combining these data sources a selection of interpolation method needs to occur. The interpolation was linear, both temporally and spatially. This might influence the results but was not considered in this study.

\begin{figure}[h]
    \caption{A flow chart showing how data sources were combined}
    \label{fig:data_preprocessing_flow_chart}
    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
\end{figure}

The procedure of combining these sources was as follows and can be seen in Figure (\ref{fig:data_preprocessing_flow_chart}). The measured data from the AWS is filtered by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed as there are the highest wind gusts. After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API. The CARRA API needs to be queried for given hours, days, months, years and a given area. That is, if queried for a given hour, it returns that hour for every day that is queried. Similarly if queried for a given day, it returns that day for every month. In light of these restraints, it was decided to query month by month. Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and interpolate for the 10 minute values. After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe. After this the downloaded data is discarded and the next month is queried. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).

Once CARRA data has been merged with AWS data, using station and time columns, then this combined file needs to checked for nails. This is done by using the hourly data (which is supposedly error free). As the data has been filtered in for the highest average wind speed in a 48 hour interval, the hourly data can be used to find nails. The hourly data is combined with the merged AWS and 10 min data. Then filtering is applied on the average wind. If the average wind speed differs the rows are dropped. This can be used to look at which stations have high percentage error. This can be seen in Figure (\ref{fig:error_count}).

\begin{figure}
    \caption{The number of stations whose number of nails are in a certain percentage range of the overall}
    \label{fig:error_count}
    \includegraphics[scale=0.75]{Figures/error_count.png}
\end{figure}

Figure (\ref{fig:error_count}) shows that most stations have nails in less that 10\% measurements. One has nails in 100\% of measurements. This is only because the number of measurements that matched previous criteria was 1.

@@

\section{Data structure}

Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test. Starting with a dataframe that contains measured information from AWS. This includes the average wind, the wind gust, wind direction along with the station number and coordinates. When selecting the CARRA data certain height levels are chosen. That is at which heights above ground the reanalysis parameters are requested. These present as separate lines in the CARRA dataframe. That is information for one observation is represented in as many lines as height levels requested in the reanalysis data. These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station. When this is done it is possible to combine the AWS Veðurstofu data and CARRA reanalysis data on the location and time columns. The last data source is the elevation. A couple of different sections of land around the weather stations have been looked at. A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point. In any case the points, that represent these sections, were selected like as shown in Code Listing (\ref{code:sectorElevation}). That is a range of angles are defined based on the wind direction d at some distance from the given point. This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.

\begin{lstlisting}[language = Python, caption = {Sector elevation points generated}, label = code:sectorElevation]
angles = [(angle + (90 - d)) * pi/180 for angle in angleRange]
length_rng = [(exp(i * log(n + 1)/ k) - 1) * 1000 
                for i in range(1, k + 1)]
points = np.array([[(X + l * cos(angle), Y + l * sin(angle))
                    for angle in angles] for l in length_rng])   
\end{lstlisting}

The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from CARRA, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data. An example of what the data looks like can be seen in Table (\ref{table:trainDataExample}).

@@

Looking at Table (\ref{table:trainDataExample}) note that the last \nPCA columns represent the principle component analysis of the elevation data. Also note that the first four columns represent two variables that describe the stability of the air. These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$). They are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}. They are calculated using reanalysis data about the the weather at two different height levels. Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1. Exactly the same notation is used with the Brunt–Väisälä frequency. The height levels used were 15, 250 and 500 meters above the ground.

>>>>

This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models. When working with hundred of thousands of rows, the efficiency of the code is very important. Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible. The three data sources were all in a different format. Measurement data from IMO was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format. To use the data to train, these three data sources needed to be combined into one file. This was done based on the measurement data from the IMO. A limit was set on the average wind speed and it was used to select measurement points. Along with the average wind speed having to be above a certain limit, to make sure that the same weather for the same location is not duplicated, only the top wind speed in any given 48 hour period is selected. That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times. The data from IMO was supplied for 10 minute increments, while CARRA data is in 3 hour intervals. This means that to use the CARRA data to predict the measured values from IMO, temporal interpolation would need to be done. Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the IMO is given at specific locations. The elevation information was given by a 20 by 20 rectangular grid that covers Iceland. When combining these data sources a selection of interpolation method needs to occur. The interpolation was linear, both temporally and spatially. This might influence the results but was not considered in this study.

\begin{figure}[h]
    \caption{A flow chart showing how data sources were combined}
    \label{fig:data_preprocessing_flow_chart}
    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart-2024-04-08.png}
\end{figure}

The procedure of combining these sources was as follows and can be seen in Figure \ref{fig:data_preprocessing_flow_chart}. The measured data from the AWS is filtered by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed as there are the highest wind gusts. After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API. The CARRA API needs to be queried for given hours, days, months, years and a given area. That is, if queried for a given hour, it returns that hour for every day that is queried. Similarly if queried for a given day, it returns that day for every month. In light of these restraints, it was decided to query month by month. Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00) and then take these values and interpolate for the 10 minute values. After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe. After this the downloaded data is discarded and the next month is queried. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).

Once CARRA data has been merged with AWS data, using station and time columns, then this combined file needs to checked for nails. This is done by using the hourly data (which is supposedly error free). As the data has been filtered in for the highest average wind speed in a 48 hour interval, the hourly data can be used to find nails. The hourly data is combined with the merged AWS and 10 min data. Then filtering is applied on the average wind. If the average wind speed differs the rows are dropped. This can be used to look at which stations have high percentage error. This can be seen in Figure \ref{fig:error_count}.

\begin{figure}
    \includegraphics[scale=0.75]{Figures/error_count.png}
    \caption{The number of stations whose number of nails are in a certain percentage range of the overall}
    \label{fig:error_count}
\end{figure}

Figure \ref{fig:error_count} shows that most stations have nails in less that 10\% measurements. One has nails in 100\% of measurements. This is only because the number of measurements that matched previous criteria was 1.

@@

\section{Data Structure}

Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test. Starting with a dataframe that contains measured information from AWS. This includes the average wind, the wind gust, wind direction along with the station number and coordinates. When selecting the CARRA data certain height levels are chosen. That is at which heights above ground the reanalysis parameters are requested. These present as separate lines in the CARRA dataframe. That is information for one observation is represented in as many lines as height levels requested in the reanalysis data. These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station. When this is done it is possible to combine the AWS IMO data and CARRA reanalysis data on the location and time columns. The last data source is the elevation. A couple of different sections of land around the weather stations have been looked at. A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point. In any case the points, that represent these sections, were selected like as shown in Code Listing \ref{code:sectorElevation}. That is a range of angles are defined based on the wind direction $d$ at some distance from the given point. This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.

\begin{lstlisting}[style = Python, caption = {Sector elevation points generated}, label = code:sectorElevation]
angles = [(angle + (90 - d)) * pi/180 for angle in angleRange]
length_rng = [(exp(i * log(n + 1)/ k) - 1) * 1000 
                for i in range(1, k + 1)]
points = np.array([[(X + l * cos(angle), Y + l * sin(angle))
                    for angle in angles] for l in length_rng])   
\end{lstlisting}

The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from CARRA, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data. An example of what the data looks like can be seen in Table \ref{table:trainDataExample}.

@@

Looking at Table \ref{table:trainDataExample} note that the last \nPCA columns represent the principle component analysis of the elevation data. Also note that the first four columns represent two variables that describe the stability of the air. These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$). They are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}. They are calculated using reanalysis data about the the weather at two different height levels. Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1. Exactly the same notation is used with the Brunt–Väisälä frequency. The height levels used were 15, 250 and 500 meters above the ground.
